---
title: "Report"
output: 
  html_document:
    code_folding: hide
---

<h1><b><center>Fastest Way to Get Around in Manhattan</center></b></h1>


<center>Keming Zhang, Junzhe Shao, Fengjia Chen, Yuchen Zheng, Kaitlyn Wang</center>  

## Motivation & Related Work
According to the INRIX scorecard, NYC is ranked No.1 for the worst traffic in the United States. In such an urban region with a dense population, people who do not wish to use personal vehicles have other transportation options such as bike, subway, bus, or taxi. Although it might seem obvious that riding a bike would not outcompete other transportations in a long travel distance at first glance, we want to examine whether this is true in Manhattan, given that its traffic is on the top list of worst traffic in the country. Inspired by the paper Hail a cab or ride a bike? A travel time comparison of taxi and bicycle-sharing systems in New York City, we are interested in whether bicycles can compete with other transportations in terms of travel time in Manhattan in 2020 to 2021 specifically. Considering the weather, rush hour, subway traffic in Manhattan, we focused on the time in a day, weekday versus weekend and seasonal variation to investigate the best time-saving way to commute all year round. Our final goal is to make an interactive map and dashboards that shows the estimated travel time using different transportations at different time periods of the year by just one click! It is expected to be a marketing tool to help those in need of public transportation in NYC to decide the most time-saving and cost-effective way to travel.  

## Questions  

The initial question that we were trying to answer was whether riding bike is faster than taking other transportions in Manhattan all year round and to identify possible factors that might contribute to traffic time difference. To examine this question, we did not just look at the travel time of each trip with each transportation, instead, we took into consideration the seasonal variation, rush hours, weekday versus weekend and weather. While we look into the dataset and explore the relation between all numbers, we hope to learn more about the characteristics through answering the following questions:  
  
- What might be the fastest transportation to choose within Manhattan?  
- What is the most popular transportation people choose within a day?  
- How would velocities of transports change according to time and taxi zone differences?
- Is it worth commuting daily driving a car compared to using public transit regardless of the money?  
- What is the distribution of public transit speed rates and how do they vary across the whole year? And what is the pattern of change during a day, a week, a season, or even the weather?  
- What is the geometric difference in choice making and distribution in various taxi zones?  
 
While we were manipulating and having deeper exploration, more useful insights, questions and conclusions came up. During this process, we organized and optimized the raw data, and transformed them into more readable explanations after appropriate processing. Meanwhile, we tried to discover and explain more based on existing data, find out the possible factors that might influence the speed, and combine them with the statistics analysis and map searching. 

## Data Source 

* [NYC Taxi Trip Record Data](https://www1.nyc.gov/site/tlc/about/tlc-trip-record-data.page) : We obtained the yellow taxi, green taxi and taxi zone lookup table from this website. The original yellow and green taxi datasets contain files from Jan. 2020 to Jul. 2021 with 33,671,276 and 749,439 observations, respectively.  

* [Dataset on Citi Bike Trip Histories](https://ride.citibikenyc.com/system-data) : Citibike dataset was obtained from the citibike website, with files from Jan. 2020 to Oct. 2021 including information on pick up locations and times, drop off locations and time, etc in Manhattan, Brooklyn, Queens and Bronx specificaly. 

* [New York City MTA](https://transitfeeds.com/p/mta) : The NYC subway schedule was obtained from the NYC MTA website, which includes 1000 trains' information on stops, arrival and departure time etc. 

* [Climate Data Online](https://www.ncdc.noaa.gov/cdo-web/) : Weather data were chosen from Jun. 2020 to May 2021 with information on months, precipitation etc.
  
## __Data Cleaning__

### General Techniques
The original dataset of taxi and bike is huge, contains ~30,000,000 each. So `dplyr` package is not suitable for that large amont of data which may be intensive for memeory and compuation. 
To handle tabular data, `data.table` is an alternative to R's default `data.frame` or `tibble`. Its popularity stems from the speed with which it can process enormous amounts of data and the brevity of its syntax. As a result, you enter less code and move at a far faster pace.

A key idea for our data cleaning is creating what we call `pickup - dropoff pairs`as a standard basic data entry. It consists of a pickup location, a drop off location, pick up time, drop off time and it's distance. The bike data can be easily transformed into pickup - dropoff pairs. The detail of transformation of MTA data will be described in the corresponding cleaning section.

### Bike Data 
Read in the bike data after sample
```{r eval=FALSE, echo = TRUE}
bike_dt = fread("citibike_manhattan_sample.csv")
bike_dt = bike_dt %>% janitor::clean_names() %>% as.data.table()
head(bike_dt)
bike_dt = bike_dt[, distance_hav := distHaversine(matrix(c(start_station_longitude, start_station_latitude), ncol = 2),
                                   matrix(c(end_station_longitude, end_station_latitude), ncol = 2))]
bike_dt = bike_dt[, distance_ew := distHaversine(matrix(c(start_station_longitude, start_station_latitude), ncol = 2),
                                                  matrix(c(end_station_longitude, start_station_latitude), ncol = 2))]
bike_dt = bike_dt[, distance_ns := distHaversine(matrix(c(end_station_longitude, start_station_latitude), ncol = 2),
                                                  matrix(c(end_station_longitude, end_station_latitude), ncol = 2))]
bike_dt = bike_dt[,angle_ns := atan(distance_ew/distance_ns)*2*(as.numeric(I(start_station_longitude < end_station_longitude))-0.5)*2*(as.numeric(I(start_station_latitude > end_station_latitude))-0.5) - manhattan_angle]
bike_dt = bike_dt[,distance := distance_hav*(abs(cos(angle_ns))+abs(sin(angle_ns)))]
colnames(bike_dt)
```
Some key variables discription:  

* `start_time`: bike start time, a POSIXct item,`end_time` similar  
* `start/end_station_lattitude/longitude`: lattitude/longitude of start/end station  
* `start/end_neighborhood`: the neighborhood, same to the taxi zone  
* `start/end_neighborhood_id`:the neighborhood id, same to the taxi zone id  

Next we need to estimate the distance.In the citibike dataset, trip distance was not provided, The main idea here is treat manhattan blocks as square in 28.9 degree north east, and caculate the angel $\theta$ between the start point-end point line and the block direction of manhattan(28.9 degree). Then the real distance$L_t$ can be estimated from the Haversine distance calculated from lattitude/longitude pairs $L_0$ by $L_t =L_0(\sin(\theta)+\cos(\theta))$.  

`distance_hav` is caculated by`distHaversine()`which is a provided function to calculate the distance from lattitude/longitude pairs from `geosphere`package. then we calculate the `distance_ew` and `distance_ns` to obtain a direction of start point-end point line. `angle_ns` is the $\theta$ between the start point-end point line and the block direction of manhattan(28.9 degree). And we finally obtain the `distance` object. 

Then we do some further data cleaning to format the bike data into our standard data format.
```{r eval=FALSE, echo = TRUE}
bike_dt_select = data.table(DOLocationID = bike_dt$start_neighborhood_id,
                            PULocationID = bike_dt$end_neighborhood_id,
                            tpep_pickup_datetime = bike_dt$start_time,
                            tpep_dropoff_datetime = bike_dt$end_time,
                            trip_distance = bike_dt$distance,
                            PUZone = bike_dt$start_neighborhood,
                            DOZone = bike_dt$end_neighborhood)
bike_dt_select$type = "bike"
```

### Taxi data

```{r eval=FALSE, echo = TRUE}
# read, clean, and sample -----------
res_dt_y = data.table(matrix( ncol = 0, nrow = 0))
#colnames(res_dt_y) = colnames(test_dt_zone)
for (x in list.files("raw_data/data/yellow_taxi")){
  temp_dt = fread(paste("raw_data/data/yellow_taxi/",x,sep =""))
  temp_dt = merge.data.table(temp_dt, PUloation_list, by.y = "LocationID", by.x = "PULocationID")
  temp_dt = merge.data.table(temp_dt, DOloation_list, by.y = "LocationID", by.x = "DOLocationID")
  temp_dt = as.data.table(sample_n(as_tibble(temp_dt),round(nrow(temp_dt)/10)))# random sample 10%
  cat(colnames(temp_dt))
  res_dt_y = rbind(res_dt_y,temp_dt, fill = T)
  print(x)
  print(nrow(temp_dt))
  print(nrow(res_dt_y))
}
# read, clean, and sample
res_dt_g = data.table(matrix(ncol = 0, nrow = 0))
#colnames(res_dt) = colnames(test_dt_zone)
for (x in list.files("raw_data/data/green_taxi")){
  temp_dt = fread(paste("raw_data/data/green_taxi/",x,sep =""))
  temp_dt = merge.data.table(temp_dt, PUloation_list, by.y = "LocationID", by.x = "PULocationID")
  temp_dt = merge.data.table(temp_dt, DOloation_list, by.y = "LocationID", by.x = "DOLocationID")
  temp_dt = as.data.table(sample_n(as_tibble(temp_dt),round(nrow(temp_dt)/10)))# random sample 10%
  #cat(colnames(temp_dt))
  res_dt_g = rbind(res_dt_g,temp_dt, fill = T)
  print(x)
  print(nrow(temp_dt))
  print(nrow(res_dt_g))
}
```
we iteratively read in all target data. Merge the taxi data with the taxi zone data list. and randomly select 10% of samples
Some key variables discription:  

* `PU/DOLocationID`: ID of pickup/drop off location  
* `tpep_pickup/dropoff_datetime`: date and time of pickup/drop off,a POSIXct item  
* `trip_distance` : travel distance in mile  

We can see we have create the basic `pickup - dropoff pairs`

### MTA 

basic analysis tools:
For the subway datasets, we were unable to obtain the real time MTA subway schedule, so the dataset was found on the New York City MTA with no data description. Besides, the subway dataset only categorizes the data to weekday, saturday and sunday, making it inconsistent with other datasets and difficult in making plots for comparison with other transportations. To solve this issue, 
we apply a random simulation with replacement to create the pickup-dropoff pair.

```{r eval=FALSE, echo = TRUE}
mta_dt  = fread('subway_schedule_clean.csv')
test_mta_dt = mta_dt[,.SD[sample(.N,5*.N, replace = T)],by = c("train_id","day","line")]
test_mta_dt = test_mta_dt[, tpep_pickup_datetime := as.POSIXct(departure_time,
                                                               format = "%H:%M:%S",
                                                               tz = "America/New_York")]
test_mta_dt_do = mta_dt[,.SD[sample(.N,5*.N, replace = T)],by = c("train_id","day","line")][, tpep_dropoff_datetime := as.POSIXct(arrival_time,format = "%H:%M:%S",tz = "America/New_York")]             
test_mta_dt= bind_cols(test_mta_dt,test_mta_dt_do) %>% tibble::repair_names()
test_mta_dt = test_mta_dt[, trip_distance := distHaversine(matrix(c(stop_lon...9, stop_lat...8), ncol = 2),
                                                  matrix(c(stop_lon...21, stop_lat...20), ncol = 2))]
test_mta_dt = test_mta_dt[,time:= abs(as.numeric(tpep_dropoff_datetime - tpep_pickup_datetime))]
test_mta_dt = test_mta_dt[,velocity := trip_distance/time]
test_mta_dt = test_mta_dt[time > 60]
test_mta_dt = test_mta_dt[velocity > 0] [velocity < 40]
setnames(test_mta_dt, "neighborhood_id...11", "PULocationID")
setnames(test_mta_dt, "neighborhood_id...23", "DOLocationID")
setnames(test_mta_dt,  "neighborhood...10" , "PUZone")
setnames(test_mta_dt,  "neighborhood...22" , "DOZone")
test_mta_dt = test_mta_dt[day...2 == "Sunday", week:=0]
test_mta_dt = test_mta_dt[day...2 == "Saturday", week:=6]
test_mta_dt[day...2 == "Weekday",]$week = sample(1:5,nrow(test_mta_dt[day...2 == "Weekday",]),replace = T)
test_mta_dt$month = sample(1:12,nrow(test_mta_dt),replace = T)
test_mta_dt = test_mta_dt[,type := "MTA"]
test_mta_final = test_mta_dt[,colnames(test_dt), with = F]
```

The key idea here is we consider all the transit without transfers between lines which the schedule can not simulate. We create 5 times random pairs grouped by `train_id`,`day`,`line` to make sure they are exactly the same train, same line, and same day. 5 fold is enough here because the variance of velocity for the same route is actually small compared to bike and taxi from our prior knowledge. 

### Intergration
New variables were added for visualization purposes in the `taxi_sample.csv`, `bike_sample.csv` and `subway_sample.csv` files. Below are the variables added:

- `Season`: winter(Dec-Feb), spring(Mar-May), summer(Jun-Aug), fall(Sep-Nov)
- `Duration`: the difference between pick up time and drop off time for citibike and subway: the time difference of a trip between a location pair for subway
- `Time of day`: rush_hour_am(7am - 10am), midday(10am - 4pm), rush_hour_pm(4pm - 7pm), evening(7pm - 12am), early_morning(12am - 7am)
- `Velocity`: trip distance divided by duration
- `Day Type`: weekday, weekend 


## Visualization

### Most Frequent Origin - Destination Pairs

Looking at the most frequent origin-destination pairs in our dataset, we can see which areas in Manhattan people prefer getting around by taxi and which area in Manhattan people prefer getting around by bike.   

### Taxi

From this barplot we can see that the top 15 most frequent origin-destination pairs are located in the upper east side and in the upper west side. 

### Bike  

From this barplot we can see that the top 15 most frequent origin-destination pairs for bike are located in East Village, East Chelsea and Central Park. 

To get a general understanding of the travel time and the travel velocity of each type of transportation, we made heatmaps and boxplots to show travel time and velocity during different time periods of the day and across seasons.

### Taxi Travel Time vs. Bike Travel Time  

These two heatmaps compare the travel time between taxi and bike. We can see that the travel time for bike are overall greater than the travel time for taxi for all day and all seasons. The travel times are the greatest during rush hour from 4pm to 7pm for both taxi and bike. Across four seasons, the travel time is the greatest in winter for taxi and the travel time is the greatest in spring for bike.  

These two heatmaps compare the velocity between taxi and bike. We can see that the velocity for bike is overall smaller than the velocity for taxi all day and all seasons. Taxi travels the slowest during rush hour from 4pm to 7pm all seasons except fall and taxi travels the slowest in winter. Similar to taxi, bike travels the slowest during rush hour from 4pm to 7pm all seasons except fall. 

This duration boxplot compares the travel time on weekdays and weekends for each type of transportation. From this plot we can see that the travel time is greater on weekend for bike. The travel time seems to be similar for subway on both weekdays and weekend. The travel time for taxi is longer on weekdays.

This velocity boxplot compares the travel velocity on weekdays and weekends for each type of transportation. From this plot we can see that the velocities for subway on both weekdays and weekend are the greatest and they are similar across the week. Taxi travels faster on the weekend and bike travels faster on weekdays.    

## Weather Influence

## Exploratory Analysis

### Analysis

Analysis methods here is simple and straightforward. We find that the velocity fits gaussian distribution well. a simple `t.test()` or `aov()` test will be good. Since we are considering a more detailed location based model. multinomial mixed logistic model which is generally the best model for transportation decision question is not necessary here. We have the following function to make the selection process tidy and we will adopt `data.table` all the time to make the deployment effcient. Here is a sample function for analysis on our website. We can analysis result from any scale simply by cancel some key variables

```{r eval=FALSE, echo = TRUE}
test_function = function(month_in, week_in, data = test_dt,PU,DO, distance_range_low = 0, distance_range_up = 6){
  x = data[month %in% month_in][week %in% week_in][type == 'taxi'][trip_distance >= 1000*distance_range_low][trip_distance <= 1000*distance_range_up][PUZone == PU][DOZone == DO][,.(velocity)]
  x %>% hist()
  y = data[month %in% month_in][week %in% week_in][type == 'bike'][trip_distance >= 1000*distance_range_low][trip_distance <= 1000*distance_range_up][PUZone == PU][DOZone == DO][,.(velocity)]
  hist(as.numeric(as.data.frame(y)))
  z = data[month %in% month_in][week %in% week_in][type == 'bike'][trip_distance >= 1000*distance_range_low][trip_distance <= 1000*distance_range_up][PUZone == PU][DOZone == DO][,.(velocity)]
  hist(as.numeric(as.data.frame(z)))
}
test_function(month_in = c(1,6,7),week_in = c(1,2,3), PU = "Washington Heights South", DO = "Morningside Heights",
              distance_range_low = 0, distance_range_up = 20, data = rbind(test_dt,test_mta_final))
```

Afterwards we can extract the mean estimate and a p.value based on either t.test or ANOVA.

### Regression 
A regression analysis on 
```{r include = FALSE}
library(tidyverse)
weather_df = read_csv('weather_transport_df.csv')
```
```{r}
lm_velocity <-
  weather_df %>%
  filter(type == "bike") %>%
  mutate(month = as.factor(month)) %>%
  drop_na()
lm(velocity ~ trip_distance + awnd + prcp + snwd + tmax + tmin, data = weather_df) %>% summary()
```

## Interaction Map
